
---
title: "Multinomial Model for ordinal data: Categorical features"
output:
  pdf_document: default
  html_document:
    highlight: pygments
    theme: spacelab
---

```{r setup, echo =FALSE, include=FALSE}
# DO NOT ALTER CODE IN THIS CHUNK
# The very first time you use this R markdown file, you should install each of the packages below.
# The same goes for other packages you might decide to use
# Remember that you only need to install each R package once in R (ever -- unless you change computers).
# All you need to do whenever you need to use the package again (after restarting the R session),
# is to use the library function to call the package.
# For example, type install.packages("knitr") in the console to install the knitr package. 
knitr::opts_chunk$set(echo = TRUE, fig.pos = 'h', fig.align = 'center')
knitr::opts_chunk$set(fig.cap = "",  fig.path = "Plot")
library(knitr)
library(dplyr)
library(arm)
library(pROC)
library(tidyverse)
library(MASS)
library(tigerstats)
library(leaps)
library(car)
library(rms)
require(caret)
require(e1071)
library(lme4) 
library(lattice)
library(broom)
library(boot)
library(ggplot2)
library(cobalt)
require(tidyverse)
require(rstanarm)
require(magrittr)
require(rstan)
require(MCMCpack)
library(abind)
library(matrixStats)
require(dae)
```

* * *

Setting: 

- For each observations, we have 3 associated variables

X1: Nomial variable with 3 possible outcomes {1,2,3} which we will generate from $Mult(1, [0.2, 0.3, 0.5])$
X2: Nomial variable with 3 possible outcomes {1,2,3} which we will generate from $Mult(1, [0.3, 0.5, 0.2])$
Y: Ordinal variable with 5 levels, {1,2,3,4,5} generated using the following process

$\epsilon_i \sim N(0,1)$
$z_i = \beta^TX_i + \epsilon$ where $X = [X1==1, X1==2, X2==1, X2==2]$ i.e. we drop category (1,1) as the baseline
$g(z_i) = Y_i$.

Here $\beta = [-3, 2, 2, -4]$ and function $g$ will be the binning function which will bin data into 5 different bins corresponding to 5 possible ordinal outcome.

- We will try to model Y conditioned on other variables (X1, and X2) using multinomial model with the parameter $\theta \in R^{45}$ governing the joint probability $P(X1, X2, Y)$. 

- We have one unknown parameter $\theta$ which we will put Dirichlet distribution prior with parameter $\alpha = 1$ as an noninformative prior.

```{r}
# Data generating process
set.seed(0)
n = 600
beta = c(-3, 2, 2, -4)

# noise term
epsilon = rnorm(n, mean = 0, sd = 1)

# X1
X1 = t(rmultinom(n, size = 1, prob = c(0.2, 0.3, 0.5)))

# X2
X2 = t(rmultinom(n, size = 1, prob = c(0.3, 0.5, 0.2)))

# X
X = cbind(X1[,2:3], X2[,2:3])
colnames(X) <- c('X1_cat2', 'X1_cat3', 'X2_cat2', 'X2_cat3')

# Z
Z = X%*%beta + epsilon

# Cut-off points and Y
g = quantile(Z, probs = c(0.2, 0.4, 0.6, 0.8))
Y = rep(NA, n)
Y[Z<g[1]] = 1
Y[Z>=g[1] & Z<g[2]] = 2
Y[Z>=g[2] & Z<g[3]] = 3
Y[Z>=g[3] & Z<g[4]] = 4
Y[Z>=g[4]] = 5
```

Model specifications:

- $\theta = (\theta_1, \theta_2, ..., \theta_{45})$ a vector of Multinomial parameter

- For all samples $x_i \sim Mult(1, \theta)$ and $x_i = (x_{1i}, x_{2i}, y_i)$

- prior distribution: $\theta \sim Dir(1)$ the non-informative prior

```{r}
# Data summary: Contingency table
df <- data.frame(cbind(apply(t(t(X1)*c(1,2,3)),1,sum), 
                       apply(t(t(X2)*c(1,2,3)),1,sum), Y))
colnames(df) = c('X1', 'X2', 'Y')
contingency_table = table(df$X1, df$X2, df$Y)

# Update posterior parameters
set.seed(1)
alpha = rep(1, 45) + matrix(contingency_table, nrow =1)

# Sample 10000 theta from posterior distribution
theta_X = rdirichlet(10000, alpha)

```

```{r}
# Imputation accuracy
empirical_pmfy = table(Y)/n
posterior_pmfy = c(mean(apply(theta_X[, 1:9], MARGIN = 1, sum)),
                  mean(apply(theta_X[, 10:18], MARGIN = 1, sum)),
                  mean(apply(theta_X[, 19:27], MARGIN = 1, sum)),
                  mean(apply(theta_X[, 28:36], MARGIN = 1, sum)),
                  mean(apply(theta_X[, 37:45], MARGIN = 1, sum)))

df3 = rbind(empirical_pmfy, posterior_pmfy)
colnames(df3)<- c('cat 1', 'cat 2', 'cat 3', 'cat 4', 'cat 5')
barplot(df3, xlab = 'Category', beside = TRUE, 
        legend = TRUE, args.legend=list(x='bottomleft'),
        main = 'Blocked Gibbs Sampling Assessment: Marginal Y pmf')
```

* * *



